Alluvial Diagrams in ggplot2

### 2018-06-02 README.rmd

```{r eval=TRUE}
titanic_wide <- data.frame(Titanic)
head(titanic_wide)
ggplot(data = titanic_wide,
       aes(axis1 = Class, axis2 = Sex, axis3 = Age,
           weight = Freq)) +
  scale_x_discrete(limits = c("Class", "Sex", "Age")) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

@@

```{r eval=TRUE}
titanic_long <- to_lodes(data.frame(Titanic),
                         key = "Demographic",
                         axes = 1:3)
head(titanic_long)
ggplot(data = titanic_long,
       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
           weight = Freq, label = stratum)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

>>>>

```{r eval=TRUE}
titanic_wide <- data.frame(Titanic)
head(titanic_wide)
ggplot(data = titanic_wide,
       aes(axis1 = Class, axis2 = Sex, axis3 = Age,
           y = Freq)) +
  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.1, .05)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum", label.strata = TRUE) +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

@@

```{r eval=TRUE}
titanic_long <- to_lodes(data.frame(Titanic),
                         key = "Demographic",
                         axes = 1:3)
head(titanic_long)
ggplot(data = titanic_long,
       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
           y = Freq, label = stratum)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  theme_minimal() +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

### 2018-06-02 inst/doc/ggalluvial.rmd

```{r example alluvial diagram using Titanic dataset, echo=FALSE}
ggplot(data = to_lodes_form(as.data.frame(Titanic),
                            key = "Demographic",
                            axes = 1:3),
       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
           weight = Freq, label = stratum)) +
  scale_x_discrete(expand = c(.05, .05)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

@@

The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": _Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column._ An additional column contains the weight of each row, e.g. the number of observational units in the cohort. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

@@

```{r alluvial diagram of UC Berkeley admissions dataset}
ggplot(as.data.frame(UCBAdmissions),
       aes(weight = Freq, axis1 = Gender, axis2 = Dept)) +
  geom_alluvium(aes(fill = Admit), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("UC Berkeley admissions and rejections, by sex and department")
```

@@

```{r parallel sets plot of Titanic dataset}
ggplot(as.data.frame(Titanic),
       aes(weight = Freq,
           axis1 = Survived, axis2 = Sex, axis3 = Class)) +
  geom_alluvium(aes(fill = Class),
                width = 0, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", label.strata = TRUE, reverse = FALSE) +
  scale_x_continuous(breaks = 1:3, labels = c("Survived", "Sex", "Class")) +
  coord_flip() +
  ggtitle("Titanic survival by class and sex")
```

@@

Weights (and weight totals) can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata are an artifact of the alluvia and often not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` will accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:

@@

```{r time series alluvia diagram of refugees dataset}
data(Refugees, package = "alluvial")
country_regions <- c(
  Afghanistan = "Middle East",
  Burundi = "Central Africa",
  `Congo DRC` = "Central Africa",
  Iraq = "Middle East",
  Myanmar = "Southeast Asia",
  Palestine = "Middle East",
  Somalia = "Horn of Africa",
  Sudan = "Central Africa",
  Syria = "Middle East",
  Vietnam = "Southeast Asia"
)
Refugees$region <- country_regions[Refugees$country]
ggplot(data = Refugees,
       aes(x = year, weight = refugees, alluvium = country)) +
  geom_alluvium(aes(fill = country, colour = country),
                alpha = .75, decreasing = FALSE) +
  scale_x_continuous(breaks = seq(2003, 2013, 2)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -30, hjust = 0)) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_color_brewer(type = "qual", palette = "Set3") +
  facet_wrap(~ region, scales = "fixed") +
  ggtitle("refugee volume by country and region of origin")
```

@@

No `weight` is specified, so each row is assigned unit weight.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

@@

```{r alluvial diagram of vaccinations dataset}
data(vaccinations)
levels(vaccinations$response) <- rev(levels(vaccinations$response))
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           weight = freq,
           fill = response, label = response)) +
  scale_x_discrete(expand = c(.05, .05)) +
  geom_flow() +
  geom_stratum(alpha = .5) +
  geom_text(stat = "stratum", size = 3) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses at three points in time")
```

@@

>>>>

```{r example alluvial diagram using Titanic dataset, echo=FALSE}
ggplot(data = to_lodes_form(as.data.frame(Titanic),
                            key = "Demographic",
                            axes = 1:3),
       aes(x = Demographic, stratum = stratum, alluvium = alluvium,
           y = Freq, label = stratum)) +
  scale_x_discrete(expand = c(.05, .05)) +
  geom_alluvium(aes(fill = Survived)) +
  geom_stratum() + geom_text(stat = "stratum") +
  ggtitle("passengers on the maiden voyage of the Titanic",
          "stratified by demographics and survival")
```

@@

The wide format reflects the visual arrangement of an alluvial diagram, but "untwisted": Each row corresponds to a cohort of observations that take a specific value at each variable, and each variable has its own column. An additional column contains the weight of each row, e.g. the number of observational units in the cohort, which may be used to control the heights of the strata.[^weight-y] Basically, the wide format consists of _one row per alluvium_. This is the format into which the base function `as.data.frame()` transforms a frequency table, for instance the 3-dimensional `UCBAdmissions` dataset:

@@

```{r alluvial diagram of UC Berkeley admissions dataset}
ggplot(as.data.frame(UCBAdmissions),
       aes(y = Freq, axis1 = Gender, axis2 = Dept)) +
  geom_alluvium(aes(fill = Admit), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Gender", "Dept"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  ggtitle("UC Berkeley admissions and rejections, by sex and department")
```

@@

```{r parallel sets plot of Titanic dataset}
ggplot(as.data.frame(Titanic),
       aes(y = Freq,
           axis1 = Survived, axis2 = Sex, axis3 = Class)) +
  geom_alluvium(aes(fill = Class),
                width = 0, knot.pos = 0, reverse = FALSE) +
  guides(fill = FALSE) +
  geom_stratum(width = 1/8, reverse = FALSE) +
  geom_text(stat = "stratum", label.strata = TRUE, reverse = FALSE) +
  scale_x_continuous(breaks = 1:3, labels = c("Survived", "Sex", "Class")) +
  coord_flip() +
  ggtitle("Titanic survival by class and sex")
```

@@

Heights can vary from axis to axis, allowing users to produce bump charts like those showcased [here](http://imgur.com/gallery/gI5p7).[^geom-area] In these cases, the strata contain no more information than the alluvia and often not plotted. For convenience, both `stat_alluvium()` and `stat_flow()` will accept arguments for `x` and `alluvium` even if none is given for `stratum`.[^arguments] As an example, we can group countries in the `Refugees` dataset by region, in order to compare refugee volumes at different scales:

@@

```{r time series alluvia diagram of refugees dataset}
data(Refugees, package = "alluvial")
country_regions <- c(
  Afghanistan = "Middle East",
  Burundi = "Central Africa",
  `Congo DRC` = "Central Africa",
  Iraq = "Middle East",
  Myanmar = "Southeast Asia",
  Palestine = "Middle East",
  Somalia = "Horn of Africa",
  Sudan = "Central Africa",
  Syria = "Middle East",
  Vietnam = "Southeast Asia"
)
Refugees$region <- country_regions[Refugees$country]
ggplot(data = Refugees,
       aes(x = year, y = refugees, alluvium = country)) +
  geom_alluvium(aes(fill = country, colour = country),
                alpha = .75, decreasing = FALSE) +
  scale_x_continuous(breaks = seq(2003, 2013, 2)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -30, hjust = 0)) +
  scale_fill_brewer(type = "qual", palette = "Set3") +
  scale_color_brewer(type = "qual", palette = "Set3") +
  facet_wrap(~ region, scales = "fixed") +
  ggtitle("refugee volume by country and region of origin")
```

@@

The stratum heights `y` are unspecified, so each row is given unit height.
This example demonstrates one way **ggalluvial** handles missing data. The alternative is to set the parameter `na.rm` to `TRUE`.[^na.rm] Missing data handling (specifically, the order of the strata) also depends on whether the `stratum` variable is character or factor/numeric.

@@

```{r alluvial diagram of vaccinations dataset}
data(vaccinations)
levels(vaccinations$response) <- rev(levels(vaccinations$response))
ggplot(vaccinations,
       aes(x = survey, stratum = response, alluvium = subject,
           y = freq,
           fill = response, label = response)) +
  scale_x_discrete(expand = c(.1, .1)) +
  geom_flow() +
  geom_stratum(alpha = .5) +
  geom_text(stat = "stratum", size = 3) +
  theme(legend.position = "none") +
  ggtitle("vaccination survey responses at three points in time")
```

@@

[^weight-y]: Previously, weights were passed to the `weight` aesthetic rather than to `y`. This prevented `scale_y_continuous()` from correctly transforming scales, and anyway it was inconsistent with the behavior of `geom_bar()`.
